DeepSphere: Efficient spherical Convolutional Neural Network with HEALPix sampling for cosmological applications

### 2019-03-04 rebuttal.tex

We performed the comparison with these features at it reflects the two-steps traditional way in classic machine learning: a) designing features b) using a method a ``simple'' classifier. Deep learning changed this tactic with networks that learn the features directly from the raw data.
The PS and histograms were chosen because they are standard statistics used by the cosmology community for analysing real data.
This is why we chose it as the reference point.
In fact, the single step of moving away from these hand-picked statistics to ML approaches is already interesting for physicists.
It would be indeed interesting to compare the performance of the CNN to other ML-based algorithms, but we belive it is a major undertaking if to be conducted fairly.
Given that the our method gives significant advantage over the traditional statistics used in cosmology, we believe that the current scope of the paper is already interesting for cosmologists.
We would be keen to leave more detailed comparisons of ML-algorithms for future work.

%In practice, those models are limited by the computational power and the dataset size.

This being said, we did try to train an SVM on the raw data, but were unable to obtain over $60\%$ accuracy in the three noiseless cases, which is far behind every other tested model.
Practically, the classifier would either overfit or be over-regularized.
We hence decided not to report the raw-data SVM results in the paper.
This information has now been added to the manuscript.
% \nati{FYI, I did try again to be entirely sure about it.}

While fully-connected NNs (FCNNs) generalize CNNs, they don't perform as well as CNNs on tasks where the symmetries exploited by the CNN are relevant.\footnote{For cosmological data on the sphere, a desired symmetry, highlighted in the introduction, is for the model to be invariant to rotations.}
That is similar to our comparison of the FCN and CNN variants.
While the CNN is a generalization, it doesn't perform as well as the FCN. We agree that show this in practice would be a great addition to the paper. Nevertheless, we were unable to make FCNN reach a baseline performance.
In practice, the large number of parameters required by a FCNN severely restricts what can be tested on conventional hardware. In our case the input number of pixels (for order $1$) is $1024^2$ which limits the number of neurons that can be used in the architecture. For completeness, we tried a FCNN and were only able to squeeze $128$ neurons in the first layer before hitting an out-of-memory error. With this architecture and a relative amount of noise of $0.5$, we were unable to regularize the network such that it would classify the validation set. (Remember that all reported classifiers obtained close to $100\%$ accuracy in this setting.) Similar conclusions were observed with other settings. These results were not included as they simply too far from the SVM classifiers. 
%Furthermore, the harmonic resemblance is another sign that the constructed graph is able to capture the spherical structure of the HEALPix sampling, even when increasing the number of neurons for small part of the sphere (order $2$ and $4$).

@@

\subsection{}
\begin{mdframed}[style=comment]
Page 4: "There is no known point set that achieves the analogue of uniform sampling in Euclidean space and allows exact and invertible discrete spherical harmonic decompositions of arbitrary but band-limited functions." This is ambiguous. There exists multiple schemes for performing exact SHT on band-limited signals. Those sampling are more efficient than healpix - often in terms of number of pixels, and sometimes in terms of performance of the spherical harmonic transforms - and more accurate since the transform is exact (at least numerically - so typical errors are machine precision). Healpix transforms are only exact to 1e-7 when using iterative schemes, and that depends on Nside and lmax. Please clarify this paragraph. Obviously healpix is very advantageous for the pooling and other operations that take advantage of its hierarchical nature. This is well discussed in the paper - in particular how the fast graph algorithms could accommodate other sampling schemes.
\end{mdframed}

That is right.
Thanks for pointing it out.
The confusion was due to our ignorance.
We removed the part about exact SHTs from the paragraph as we don't need to make such a point.
%\mdeff{That sentence mostly came from \url{https://healpix.jpl.nasa.gov/html/intronode2.htm}.}
%\mdeff{uniform sampling == same distance between all neighboring pixels}

\subsection{}
\begin{mdframed}[style=comment]
"Our graph is constructed as an approximation of sphere S2, a 2D manifold embedded in R3, Indeed, [33] showed that the graph Laplacian converges to the Laplace-Beltrami when the number of pixels goes to infinity. While our construction will not exactly respect the setting defined by [33], we observe empirically strong evidence of convergence." This paragraph is unclear. Could you please add a few words to clarify what the approach of 33 is?
\end{mdframed}

We tried to clarify the paragraph in the following manner:
\begin{mdframed}[style=manuscript]
Our graph is constructed as an approximation of the sphere $S^2$, a 2D manifold embedded in $\mathbb{R}^3$.
Indeed, \citep{belkin2007convergence} showed that the graph Laplacian converges to the Laplace-Beltrami when the number of pixels goes to infinity providing uniform sampling of the manifold and a fully connected graph built with exponentially decaying weights.
While our construction does not exactly respect their setting (the sampling is deterministic and the graph is not fully connected), we empirically observe a strong correspondence between the eigenmodes of both Laplacians (see Appendix A).
\end{mdframed}

@@

Those maps have been produced with a resolution of $N_{side}=16$. The information has been added to the paper. (High resolution is costly as diagonalizing the Laplacian matrix scales as $O(N_{pix}^3)$.)

Since we do not have any convergence result yet, we are not sure about how the projection error evolves. However, if there is convergence (and we believe there is providing the correct set of graph weight is selected), the eigenvectors associated with the lowest $\ell$ probably converge first.

\subsection{}
\begin{mdframed}[style=comment]
Page 7: "That is much more efficient than filtering with spherical harmonics, even though HEALPix was de- signed as an iso-latitude sampling that has a fast spherical transform. That is especially true for smooth kernels which require a low polynomial degree K. Figure 4 compares the speed of low-pass filtering for Gaussian smoothing using the spherical harmonics and the graph-based method presented here. A naive implementation of our method is ten to twenty times faster for Nside = 2048, with K = 20 and K = 5, respectively, than using the spherical harmonic transform at lmax = 3Nside implemented by the highly optimized libsharp [38] library used by HEALPix." This is not so clear. Is the graph method also performing a Gaussian smoothing with the same characteristics? If so, what are those? How precise is the graph convolution approximating the true harmonic-space Gaussian convolution in this case? This is relevant since some approximations (for instance when analyzing cosmic microwave background data) require high-precision Gaussian smoothing. Finally, is lipsharp optimizing Gaussian smoothing in any way?
\end{mdframed}

The experiment whose result is reported in Figure 4 is intended to report filtering speed, independently of the chosen filter (we used a Gaussian for convenience).
While libsharp is probably not optimizing Gaussian smoothing in particular (w.r.t. filtering with any other filter), it is a highly optimized library for SHT.
We don't care about the approximation quality as, in the context of neural networks, filters are learned.
That is, graph filters are not approximating any predefined true harmonic-space filters anyway.
It is, however, true that replacing spherical filtering (using the SHT) with graph filtering is an interesting prospect (in terms of speed) that might be possible in the future as suggested by Appendix B (Figure B.13).
We feel that, in its current state, graph filtering is probably not a good enough approximation to be usable when high-precision filtering is required.
Nevertheless, theoretical improvements on a potential equivalence (or convergence) of the Laplacian eigenvectors to the spherical harmonics (see our answer to the comment 1.4) could potentially make it competitive.

@@

\mdeff{measure the computational time of filtering on a 1/12, 1/48 or 1/192 of the sphere with the graph. Add the result to Figure 4 and compare with filtering of the full sphere (with SHT and graph).}

\subsection{}

\begin{mdframed}[style=comment]
Secondly, using only SVM as a comparison might skew how well this technique appears to works in a deep learning setting. In principle (and I say this not knowing for sure) 2D CNNs should probably be able to work as well as DeepSphere (although no where near as elegantly) by taking in a projected 2D image of the sky and using *loads* of filters to learn about the distortion at different parts of the projected sky. It would be interesting (and I think a fairly easy test) to check how well DeepSphere performs against a 2D CNN on projections.
\end{mdframed}

%\todo{Assigned: @Nati, @Michael, @Tomek}
%\nati{Please check if you like what I did in light of the discussion we had.}

We share the opinion of the reviewer when he says that 2D convolution neural networks (2D ConvNet) should probably work as well as DeepSphere ``by taking in a projected 2D image of the sky and using *loads* of filters to learn about the distortion at different parts of the projected sky.'' Hence we decided to add a new experiment in the paper.

Nevertheless, making a comparison is technically not as trivial as it may seems.
Problematically a radial projection is computationally expensive to the point where, because of the dataset augmentation, it becomes significantly more expensive than the actual training of the network.
Hence, we decided to use the geometrical properties of HEALPix to create a ``special'' projection.
In the HEALPix sampling, the sphere is first separated into $12$ parts that are then subdivided into $4$ recursively.
Eventually, each of these $12$ parts resembles a square grid.
As the problem we are solving only involve one of these 12 parts, we can use this natural structure to simply re-assign each HEALPix pixel to one of a 2 dimensional grid of size $N_{sides} \times N_{sides}$.
This operation is computationally very cheap and probably spread the projection error relatively evenly across the resulting image.
We note that, this projection has recently been used for a CNN in cosmology in the very recent work \cite{krachmalnicoff2019convolutional}.
%As we believe that this projection is the optimal one to use a traditional CNN for applications, we performed the experiment using it.

Our expectation were alligned with the one of the reviewer, but the results did surprise us as the performance of the traditional CNN did in general not match DeepSphere, (see Figure 9 of the manuscript).
It can probably be explained by the fact that 2DCNN breaks the rotational invariance of the data.
The problem comes from the fact that the projection introduce a distortion in the geometry and the network has to learn a correct weighting scheme to compensate for it.
Unfortunately, this operation waste many parameters, reducing the modelization power of the network and requires large amount of training data which is not fully available.
Note that, when visualizing the learned convolution kernels, we observed that they were not radial.
We did not use *loads* of filters as suggested by the reviewer as the optimization was challenging and we already had to regularize significantly the network. 
While adding more filters, did improve slightly the performance (never to the level of DeepSphere), it made the optimization even more challenging to the point where we could not obtain consistent results if we were to restart the training process.
Eventually, we decided to report the results for the architecture with the same number of parameters as DeepSphere.

### 2019-03-04 scnn.tex
last pass upto comment 2.2

Our graph is constructed as an approximation of the sphere $S^2$, a 2D manifold embedded in $\mathbb{R}^3$.
Indeed, \citep{belkin2007convergence} showed that the graph Laplacian converges to the Laplace-Beltrami when the number of pixels goes to infinity providing random uniform sampling of the manifold and a fully connected graph built with exponentially decaying weights.
While our construction does not exactly respect their setting (the sampling is deterministic and the graph is not fully connected), we empirically observe a strong correspondence between the eigenmodes of both Laplacians (see \ref{sec:comparison_spherical_harmonics}).

@@

\nati{
To our surprise, the traditional 2DCNN did not match the performance of DeepSphere.
It can probably be explained by the fact that 2DCNN breaks the rotational invariance of the data.
The problem comes from the fact that the projection introduce a distortion in the geometry and the network has to learn a correct weighting scheme to compensate for it.
Unfortunately, this operation waste many parameters, reducing the modelization power of the network and requires large amount of training data which is not fully available.
Note that, when visualizing the learned convolution kernels, we observed that they were not radial.
}

@@

\nati{
\section{Maps projection onto a 2-dimensional plane}
\label{sec:projection}
In order to compare the DeepSphere architecture with a traditional 2d-CNN, we need to project the spherical data onto a plane.
Problematically a radial projection is computationally expensive to the point where, because of the dataset augmentation, it becomes significantly more expensive than the actual training of the network.
Hence, we decided to use the geometrical properties of HEALPix to create a ``special'' projection.
In the HEALPix sampling, the sphere is first separated into $12$ parts that are then subdivided into $4$ recursively.
Eventually, each of these $12$ part resembles a square grid.
To visually see it, we refer the reader to \figref{border_effects}.
As the problem we are solving only involve one of these 12 parts, we can use this natural structure to simply re-assign each HEALPix pixel to one of a 2 dimensional grid of size $N_{sides} \times N_{sides}$.
This operation is computationally very cheap and probably spread quite evenly across the resulting image.
This projection has recently been used for a CNN in cosmology in the very recent work \cite{krachmalnicoff2019convolutional}.
}

>>>>

Our graph is constructed as an approximation of the sphere $S^2$, a 2D manifold embedded in $\mathbb{R}^3$.
Indeed, \citep{belkin2007convergence} showed that the graph Laplacian converges to the Laplace-Beltrami when the number of pixels goes to infinity providing uniform sampling of the manifold and a fully connected graph built with exponentially decaying weights.
While our construction does not exactly respect their setting (the sampling is deterministic and the graph is not fully connected), we empirically observe a strong correspondence between the eigenmodes of both Laplacians (see \ref{sec:comparison_spherical_harmonics}).

@@

\nati{
To our surprise, the traditional 2D ConvNet did not match the performance of DeepSphere.
It can probably be explained by the fact that the 2D ConvNet breaks the rotational invariance of the data.
The problem comes from the fact that the projection introduce a distortion in the geometry and the network has to learn a correct weighting scheme to compensate for it.
Unfortunately, this operation waste many parameters, reducing the modelization power of the network and requires large amount of training data which is not fully available.
Note that, when visualizing the learned convolution kernels, we observed that they were not radial.
}

@@

\nati{
\section{Maps projection onto a 2-dimensional plane}
\label{sec:projection}
In order to compare the DeepSphere architecture with a 2D ConvNet, we need to project the spherical data onto a plane.
Problematically a radial projection is computationally expensive to the point where, because of the dataset augmentation, it becomes significantly more expensive than the actual training of the network.
Hence, we decided to use the geometrical properties of HEALPix to create a ``special'' projection.
In the HEALPix sampling, the sphere is first separated into $12$ parts that are then subdivided into $4$ recursively.
Eventually, each of these $12$ parts resembles a square grid.
To visually see it, we refer the reader to \figref{border_effects}.
As the problem we are solving only involve one of these $12$ parts, we can use this natural structure to simply re-assign each HEALPix pixel to one of a 2 dimensional grid of size $N_{sides} \times N_{sides}$.
This operation is computationally very cheap and probably spread quite evenly across the resulting image.
This projection has recently been used for a CNN in cosmology in the very recent work \cite{krachmalnicoff2019convolutional}.
}
